---
title: "The evolution of island geographies in Australia"
description: |
  The combination of bathymetric data and sea level curves reveals changing island geographies through the Pleistocene and Holocene.
author:
  - name: Patrick Morrison
    url: https://padmorrison.com
    affiliation: Centre for Rock Art Research + Management, University of Western Australia
    affiliation_url: https://www.crarm.uwa.edu.au/
    orcid_id: 0000-0003-0920-3405
date: 2023-02-06
output:
  distill::distill_article:
    self_contained: true
    code_folding: true
    toc: true
bibliography: inputs/references.bib
draft: false
---

This notebook contains the analysis run for 'The evolution of island geographies and the emergence and persistence of Indigenous maritime cultures' (Morrison et al. 2023.)

It relies on the sf, terra, and stars libraries. 

```{r message=FALSE, warning=FALSE, results = 'hide'}
library(terra)
library(stars)
library(tidyverse)
library(sf)
library(units)
library(lwgeom)
library(geojsonsf)
library(leaflet)
library(glue)
library(readxl)
library(patchwork)
theme_set(theme_bw())

dir.create("output_islands")
dir.create("output_islands/spatial")
dir.create("output_islands/intervis")
dir.create("output_islands/area")
dir.create("output_islands/data")
dir.create("output_islands/pub")
```

The Geoscience Australia Australian Bathymetry and Topography Grid has elevation data for the entire Australian continent at a 250m horizontal resolution. It is available as a [1.3Gb \*.ers grid file](http://pid.geoscience.gov.au/dataset/ga/67703).

The study area is defined as everything within 200km of the LGM shoreline, in an Australian Albers projection.

```{r import, message=FALSE, warning=FALSE}
#Load ausbath dataset https://doi.org/10.25919/cm17-xc81
bath <- rast('external_data/ausbath/ausbath_09_v4_ex_ex.ers') %>%
  #aggregate(2) %>% 
  terra::project('epsg:3577')

lake_eyre <- geojson_sf('inputs/lake_eyre.geojson') %>% st_transform(3577)
analysis_regions <- geojson_sf('inputs/regions.geojson') %>% st_transform(3577)


#define the study area as everything up to 200km from the lgm shorelines,
#exclude lake eyre
study_area <- st_as_stars(bath>-130) %>%
  st_as_sf(merge=TRUE) %>% 
  filter(across(1, ~ . == 1)) %>%  
  st_cast("POLYGON") %>% st_make_valid() %>% 
  mutate(area = as.numeric(set_units(st_area(.), ha))) %>% 
  filter(area>10e6) %>% 
  st_buffer(2e5) %>%
  st_difference(lake_eyre)
```

## Sea Level Figure

The sea level curve for this analysis was the Red Sea sea level record: https://doi.org/10.1038/nature11593 [@grant-2012].
Their supplementary material includes the full data, which was resampled by 1ka.

```{r fig.height=3.5, fig.width=8}
grant2012 <- read_excel('external_data/grant2012.xls',
                        sheet = "(4) RSL", range = "E9:J1210")[-1,]

sealevel <- grant2012 %>% 
  select(
    age = Age...1,
    smooth = RSL_smooth
  ) %>% mutate_all(as.double) %>% filter(age<=125)

ci <- grant2012 %>% select(
  ci_age = Age...4,
  upper = `RSL_95%upper`,
  lower = `RSL_95%lower`
) %>% mutate_all(as.double) %>% filter(ci_age<=125)

years <- seq(0,
             65,
             by = 1) #in ka

sea_spline <- with(sealevel, smooth.spline(age, smooth, all.knots=TRUE))

predictions <- predict(sea_spline, years) %>% 
  as_tibble() %>%
  dplyr::select('age' = x, "smooth"=y) %>% 
  mutate(smooth=replace(smooth, age==0, 0))


ggplot() +
    geom_ribbon(data=ci,
                aes(x = ci_age, ymin=lower, ymax=upper),
                fill='lightgrey') +
    geom_line(data=sealevel, aes(age,smooth), colour="blue4", linewidth=1) + 
    scale_x_reverse(n.breaks=12, expand = c(0, 0)) +
    scale_y_continuous(breaks=c(20,0,-40,-80,-120), position = "right") +
  annotate("text", x = 65, y = -20, label = "Human arrival\n in Sahul") +
  annotate("text", x = 24, y = -20, label = "LGM") +
    geom_vline(xintercept = c(19, 29,55,75), linetype = "dotted") +
    labs(title="Sea level since the last interglacial", subtitle = "",
         x='Thousands of years ago', y='Sea level below present (m)')

ggsave('output_islands/pub/sea_levels.png', bg='white', width = 8, height = 3.5)
```


## Islands under the sea

Then we defined a function that takes this bathymetry and sea level, and returns a vectorised landmass with area and intervisibility zones (following Kealy et al 2017).
```{r message=FALSE, warning=FALSE}
#function that takes a bathymetry and sea level, and returns the landmass
flood <- function(bath, sea_level)   {
  
  land <- st_as_stars(bath > sea_level) %>%
    st_as_sf(merge=TRUE) %>% 
    filter(across(1, ~ . == 1)) %>%  
    st_cast("POLYGON") %>% st_make_valid() %>% 
    st_filter(study_area) %>% 
    mutate(area = st_area(.),
           area = as.numeric(round(set_units(area, ha))),
           per = st_perimeter(.),
           per = as.numeric(round(set_units(per, m))),
           elev = terra::extract(x = bath, y = terra::vect(.),
                        fun=max, na.rm=TRUE)[,2]-sea_level,
           vis = (3.56*sqrt(0.5)) + 3.57*sqrt(elev))
  return(land)
}
```

Another function takes this and produces a summary of measurements for plotting, according to regions defined in the paper.
```{r}
region_calculations <- function(islands_measures, ri) {
  
  islands_measures <- filter(islands_measures, mainland_vis != 'none')
    
  island_gridded <- islands_measures %>% 
    st_join(st_set_crs(analysis_regions, st_crs(islands_measures)),
            largest=TRUE) %>% 
    group_by(region) %>% 
    summarise(
      nislands = n(),
      total_area = sum(area),
      total_per = sum(per)
      ) %>% 
    st_set_geometry(NULL)
  
  ri <- filter(ri, mainland != 'none')
  
    intervis_gridded <- ri %>% 
    st_join(st_set_crs(analysis_regions, st_crs(ri)),
            largest=TRUE) %>% 
    group_by(region) %>% 
    summarise(
      n_areas = n(),
      total_ri_area = sum(area),
      ) %>% 
    st_set_geometry(NULL)
    
    counts <- island_gridded %>% 
      left_join(intervis_gridded, by = 'region')
    
  
  
  return(counts)
}
```

Island measurements are summarised using 200km wide hexagonal bins.
```{r}
grid_islands <- function(islands_measures) {
    
  island_gridded <- islands_measures %>% 
    filter(mainland_vis != 'none') %>% 
    st_join(st_set_crs(grid_empty, st_crs(islands_measures)),
            largest=TRUE) %>% 
    group_by(grid_id) %>% 
    summarise(
      nislands = n(),
      total_area = sum(area),
      total_per = sum(per)
      ) %>% 
    filter(nislands>0) %>%
    st_set_geometry(NULL)
  
  grid <- grid_empty %>% 
    st_set_crs(st_crs(islands_measures)) %>% 
    right_join(island_gridded)
  
  return(grid)
}

#create hexagonal grid
grid_empty <- ext(bath) %>%
  as.polygons() %>%
  st_as_sf() %>% 
  st_make_grid(cellsize = 200000,
               what = "polygons", square = FALSE) %>% 
  st_sf() %>% 
  mutate(grid_id = row_number())
```

This function combines those above, and produces outputs for each sea level in the publication.
```{r}
plots_for_pub <- function(bath, sea_level) {
  
  flooded_test <- flood(bath, sea_level)
  mainland_test <- filter(flooded_test, area > 10e5)
  islands_test <- filter(flooded_test, area < 10e5)
  
  ri <- st_buffer(islands_test, islands_test$vis*1000) %>%
  st_union() %>% st_cast('POLYGON') %>% st_as_sf() %>% mutate(
     mainland = ifelse(lengths(st_intersects(., mainland_test))>0, 'mainland', 'none'),
     area = st_area(.),
           area = as.numeric(round(set_units(area, ha))))
  
  ra <- st_buffer(islands_test, (islands_test$vis/2)*1000) %>%
  st_union() %>% st_cast('POLYGON') %>% st_as_sf() %>% mutate(
     mainland = ifelse(lengths(st_intersects(., mainland_test))>0, 'mainland', 'none'),
     area = st_area(.),
           area = as.numeric(round(set_units(area, ha))))
  
  islands_test <- islands_test %>% mutate(
  mainland_vis = case_when(
    lengths(st_intersects(.,
                          filter(ra, mainland =='mainland'))) > 0 ~ 'ra',
    lengths(st_intersects(.,
                          filter(ri, mainland =='mainland'))) > 0 ~ 'ri',
    TRUE ~ 'none' 
  )
  )
  
  dir.create(glue("output_islands/pub/{sea_level}m"))
  st_write(ri, glue("output_islands/pub/{sea_level}m/ri_{sea_level}m.geojson"), append=FALSE)
  st_write(ra, glue("output_islands/pub/{sea_level}m/ra_{sea_level}m.geojson"), append=FALSE)
  st_write(islands_test, glue("output_islands/pub/{sea_level}m/islands_{sea_level}m.geojson"), append=FALSE)
  st_write(mainland_test, glue("output_islands/pub/{sea_level}m/mainland_{sea_level}m.geojson"), append=FALSE)
  
  
ggplot() +
  geom_sf(data=filter(ri, mainland=='mainland'), linewidth = 0, fill='#bcd2ee') +
  geom_sf(data=filter(ra, mainland=='mainland'), linewidth = 0, fill='#6e7b8b') +
  geom_sf(data=mainland_test, linewidth = 0.1, color='#e0e0e0') +
  geom_sf(data = islands_test, linewidth = 0.1, fill='#C4C4C4') +
  geom_sf(data = filter(islands_test, mainland_vis!='none'), linewidth = 0.1, fill='#701C00') +
  labs(title = paste0("Intervisibility surfaces"),
       subtitle = paste0(round(sea_level), "m sea level, compared to present")) +
  theme(legend.position = "bottom") +
   coord_sf(xlim = c(-2143673, 2719484), ylim = c(-5175727, -835297), expand = FALSE)
ggsave(glue("output_islands/pub/{sea_level}m/intervis_{sea_level}m.png"), bg='white', width = 7, height = 7)
ggsave(glue("output_islands/pub/{sea_level}m/intervis_{sea_level}m.pdf"), bg='white', width = 7, height = 7)

#grid the islands computed before
grid <- grid_islands(islands_test)

ggplot(grid) +
  geom_sf(data=mainland_test, linewidth = 0.1, color='#e0e0e0') +
  geom_sf(data=islands_test, linewidth = 0.1, fill='#701C00') +
  geom_sf(aes(fill=log(total_area),
              alpha =log(total_area)), color='transparent') +
  scale_fill_gradient(low = "white", high = "#015488",
                      limits=c(0,15),guide = 'none') +
  scale_alpha(guide = 'none', range = c(0.01, 0.8)) +
  #geom_sf_text(aes(label=nislands), colour='white') +
  labs(title = paste0("Total island area"),
       subtitle = paste0(round(sea_level), "m sea level, compared to present"),
       fill='log total area (ha)') +
   coord_sf(xlim = c(-2143673, 2719484), ylim = c(-5175727, -835297), expand = FALSE)
  theme(legend.position = "bottom")
  
ggsave(glue("output_islands/pub/{sea_level}m/area_{sea_level}m.png"), bg='white', width = 7, height = 7)
ggsave(glue("output_islands/pub/{sea_level}m/area_{sea_level}m.pdf"), bg='white', width = 7, height = 7)

regions_counted <- region_calculations(islands_test, ri) %>% 
  mutate(sea_level = sea_level)
write_csv(regions_counted, glue("output_islands/pub/{sea_level}m/island_counts_{sea_level}m.csv"))

}

for(sea_level in c(0,-40,-80,-120)) {
  plots_for_pub(bath = bath, sea_level = sea_level)
}

dir.create(glue("output_islands/pub/supporting"))

data <- tribble(
  ~vis, ~area,
  "Island",   0,
  "Mainland",   0,
  "Absolute Intervisibility",   15,
  "Relative Intervisibility",   15,
)

area_legend <- ggplot(data) +
  geom_col(aes(vis, area, fill = area)) +
  labs(fill = 'log(area)') +
  theme(legend.position="bottom")

ggsave("output_islands/pub/supporting/legend_area.pdf", area_legend)

intervis_legend <- ggplot(data) +
  geom_col(aes(vis, area, fill = vis)) +
  labs(fill = 'Category') +
  scale_fill_manual(values = c('Mainland' = '#E5E5E5', 'Island' = '#701C00', 'Absolute Intervisibility' = '#6e7b8b', 'Relative Intervisibility' = '#bcd2ee')) + 
  theme(legend.position="bottom")
ggsave("output_islands/pub/supporting/legend_vis.pdf", intervis_legend)
```

Similar function to above, but this one calculates it based on the year, taken from the sea level curve. Then it outputs the results for every 2,000 years.
```{r}
plot_intervis_year <- function(bath, step) {
  
  sea_level <- predictions$smooth[step]
  age <- predictions$age[step]
  
  flooded_test <- flood(bath, sea_level)
  mainland_test <- filter(flooded_test, area > 10e5)
  islands_test <- filter(flooded_test, area < 10e5)
  
  ri <- st_buffer(islands_test, islands_test$vis*1000) %>%
  st_union() %>% st_cast('POLYGON') %>% st_as_sf() %>% mutate(
     mainland = ifelse(lengths(st_intersects(., mainland_test))>0, 'mainland', 'none'),
     area = st_area(.),
           area = as.numeric(round(set_units(area, ha))))
  
  ra <- st_buffer(islands_test, (islands_test$vis/2)*1000) %>%
  st_union() %>% st_cast('POLYGON') %>% st_as_sf() %>% mutate(
     mainland = ifelse(lengths(st_intersects(., mainland_test))>0, 'mainland', 'none'),
     area = st_area(.),
           area = as.numeric(round(set_units(area, ha))))
  
  islands_test <- islands_test %>% mutate(
  mainland_vis = case_when(
    lengths(st_intersects(.,
                          filter(ra, mainland =='mainland'))) > 0 ~ 'ra',
    lengths(st_intersects(.,
                          filter(ri, mainland =='mainland'))) > 0 ~ 'ri',
    TRUE ~ 'none' 
  )
  )
  
  dir.create(glue("output_islands/spatial/{age}ka"))
  st_write(ri, glue("output_islands/spatial/{age}ka/ri_{age}ka.geojson"), append=FALSE)
  st_write(ra, glue("output_islands/spatial/{age}ka/ra_{age}ka.geojson"), append=FALSE)
  st_write(islands_test, glue("output_islands/spatial/{age}ka/islands_{age}ka.geojson"), append=FALSE)
  st_write(mainland_test, glue("output_islands/spatial/{age}ka/mainland{age}ka.geojson"), append=FALSE)
  
  
ggplot() +
  geom_sf(data=filter(ri, mainland=='mainland'), size = 0, fill='#bcd2ee') +
  geom_sf(data=filter(ra, mainland=='mainland'), size = 0, fill='#6e7b8b') +
  geom_sf(data=mainland_test, linewidth = 0.1, color='#e0e0e0') +
  geom_sf(data = islands_test, linewidth = 0.1, fill='#C4C4C4') +
  geom_sf(data = filter(islands_test,mainland_vis!='none'), size = 0.1, fill='#701C00') +
  labs(title = paste0("Intervisibility surfaces at ", age, "ka"),
       subtitle = paste0(round(sea_level), "m sea level, compared to present")) +
  theme(legend.position = "bottom") +
   coord_sf(xlim = c(-2143673, 2889484), ylim = c(-5175727, -635297), expand = FALSE)
ggsave(glue("output_islands/intervis/intervis_{age}ka.png"), bg='white', width = 7, height = 7)

#grid the islands computed before
grid <- grid_islands(islands_test)

ggplot(grid) +
  geom_sf(data=mainland_test, size = 0.1, color='#e0e0e0') +
  geom_sf(data=islands_test, size = 0.1, fill='#701C00') +
  geom_sf(aes(fill=log(total_area),
              alpha =log(total_area)), color='transparent') +
  scale_fill_gradient(low = "white", high = "#015488",
                      limits=c(0,15),guide = 'none') +
  scale_alpha(guide = 'none', range = c(0.01, 0.8)) +
  #geom_sf_text(aes(label=nislands), colour='white') +
  labs(title = paste0("Total island area at ", age, "ka"),
       subtitle = paste0(round(sea_level), "m sea level, compared to present"),
       fill='log total area (ha)') +
   coord_sf(xlim = c(-2143673, 2889484), ylim = c(-5175727, -635297), expand = FALSE)
  theme(legend.position = "bottom")
  
ggsave(glue("output_islands/area/area_{age}ka.png"), bg='white', width = 7, height = 7)


regions_counted <- region_calculations(islands_test, ri) %>% 
  mutate(sea_level = sea_level,
         age = age)
write_csv(regions_counted, glue("output_islands/data/island_counts_{age}ka.csv"))

}

for(step in seq(1,nrow(predictions),2)) {
  plot_intervis_year(bath = bath, step = step)
}
```

Finally, use the computed regional summaries to visualise a timeseries.
```{r fig.height=6, fig.width=10}
order <- c('Pilbara/Canarvon', 'Kimberley', 'Tiwi-Coburg', 'Gulf of Carpentaria', 
           'Torres Strait', 'Great Barrier Reef', 'Sydney Area', 'Bass Strait',
           'Southern Tasmania', 'Kangaroo Island Area')

region_counts <- read_csv(fs::dir_ls('output_islands/data/', regexp = "\\.csv$")) %>% 
  mutate(region = fct_relevel(region, order),
         total_area = log(total_area/100 + 1),
         total_ri_area = total_ri_area/100) %>% 
  filter(region != 'NA') %>% 
  filter((age %% 2) == 0)

ggplot(region_counts) +
  geom_line(aes(age,total_area)) +
  geom_point(aes(age,total_area), size =.5, alpha=0.3) +
  scale_y_log10() +
 facet_grid(rows = vars(region), scales = "free_y") +  ylim(c(0, NA)) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(title = 'Total Island Area Available',
       x = 'Age (ka)', y='Area log sqkm') + geom_vline(xintercept = 7) +
ggplot(region_counts) +
  geom_line(aes(age,total_ri_area)) +
  geom_point(aes(age,total_ri_area), size =.5, alpha=0.3) +
  facet_grid(rows = vars(region), scales = "free_y") +  ylim(c(0, NA)) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(title = 'Relative Intervisibility Area',
       x = 'Age (ka)', y='Area sqkm') +
  geom_vline(xintercept = 7)

ggsave('output_islands/pub/timeseries.png', bg='white', width = 14, height = 14)
```


